---
title: "Use ecocomDP Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use ecocomDP Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

A set of "use" functions help you search, read, manipulate, plot, and save ecocomDP data.

```{r, eval=FALSE, setup}
library(ecocomDP)
```

```{r include=FALSE}
library(ecocomDP)
```

## Search

Find data of interest by searching on one or more query parameters. 

All available datasets are returned with an empty call to `search_data()`:

```{r eval=FALSE}
search_data()
#> # A tibble: 82 x 11
#>   source id     title  description abstract  years sampling_interv~ sites url   
#>   <chr>  <chr>  <chr>  <chr>       <chr>     <chr> <chr>            <chr> <chr> 
#> 1 EDI    edi.3~ Fores~ <NA>        "This da~ 8     0.01             <NA>  https~
#> 2 EDI    edi.2~ Zoopl~ <NA>        "This da~ 9     0.19             <NA>  https~
#> 3 EDI    edi.3~ Long ~ <NA>        "This da~ 21    0.02             <NA>  https~
#> 4 EDI    edi.3~ Dune ~ <NA>        "This da~ 7     0.15             <NA>  https~
#> # ... with 78 more rows, and 2 more variables: source_id <chr>,
#> #   source_id_url <chr>
```

The "taxa" parameter enables searching on any taxonomic rank value. Find datasets about insects:

```{r eval=FALSE}
search_data(taxa = "Insecta")
#> # A tibble: 13 x 11
#>   source id     title  description abstract  years sampling_interv~ sites url   
#>   <chr>  <chr>  <chr>  <chr>       <chr>     <chr> <chr>            <chr> <chr> 
#> 1 EDI    edi.2~ Moss ~ <NA>        "This da~ 23    821.61           <NA>  https~
#> 2 EDI    edi.2~ Bonan~ <NA>        "This da~ 38    0.41             <NA>  https~
#> 3 EDI    edi.3~ Aphid~ <NA>        "This da~ 4     0                <NA>  https~
#> 4 EDI    edi.3~ CGR02~ <NA>        "This da~ 30    0.06             <NA>  https~
#> # ... with 9 more rows, and 2 more variables: source_id <chr>,
#> #   source_id_url <chr>
```

The "num_years" parameter lets you find datasets within a range of sampling years. Datasets with 20 to 40 years of sampling:

```{r eval=FALSE}
search_data(num_years = c(20, 40))
#> # A tibble: 22 x 11
#>   source id     title  description abstract  years sampling_interv~ sites url   
#>   <chr>  <chr>  <chr>  <lgl>       <chr>     <int>            <dbl> <chr> <chr> 
#> 1 EDI    edi.3~ Long ~ NA          "This da~    21             0.02 <NA>  https~
#> 2 EDI    edi.2~ Moss ~ NA          "This da~    23           822.   <NA>  https~
#> 3 EDI    edi.3~ Annua~ NA          "This da~    20             0    <NA>  https~
#> 4 EDI    edi.3~ SGS-L~ NA          "This da~    26             0.05 <NA>  https~
#> # ... with 18 more rows, and 2 more variables: source_id <chr>,
#> #   source_id_url <chr>

```

The "area" parameter facilitates searches by geographic regions. Datasets originating from the Northeast United States:

```{r eval=FALSE}
search_data(area = c(47, -70, 38, -90))
#> # A tibble: 55 x 11
#>   source id     title  description abstract  years sampling_interv~ sites url   
#>   <chr>  <chr>  <chr>  <chr>       <chr>     <chr> <chr>            <chr> <chr> 
#> 1 EDI    edi.3~ Fores~ <NA>        "This da~ 8     0.01             <NA>  https~
#> 2 EDI    edi.3~ Long ~ <NA>        "This da~ 21    0.02             <NA>  https~
#> 3 EDI    edi.3~ Dune ~ <NA>        "This da~ 7     0.15             <NA>  https~
#> 4 EDI    edi.3~ SGS-L~ <NA>        "This da~ 7     0                <NA>  https~
#> # ... with 51 more rows, and 2 more variables: source_id <chr>,
#> #   source_id_url <chr>
```

Combine query parameters to narrow results:

```{r eval=FALSE}
search_data(taxa = "Insecta", num_years = c(20, 100), area = c(47, -70, 38, -90))
#> # A tibble: 4 x 11
#>   source id     title  description abstract  years sampling_interv~ sites url   
#>   <chr>  <chr>  <chr>  <lgl>       <chr>     <int>            <dbl> <chr> <chr> 
#> 1 EDI    edi.3~ CGR02~ NA          "This da~    30             0.06 <NA>  https~
#> 2 EDI    edi.2~ North~ NA          "This da~    36            10.5  <NA>  https~
#> 3 EDI    edi.2~ North~ NA          "This da~    35            10.8  <NA>  https~
#> 4 EDI    edi.2~ North~ NA          "This da~    36            10.5  <NA>  https~
#> # ... with 2 more variables: source_id <chr>, source_id_url <chr>
```

Assign results to a variable and learn more:

```{r eval=FALSE}
r <- search_data(taxa = "Insecta", num_years = c(20, 100), area = c(47, -70, 38, -90))
r$title
#> [1] "CGR02 Sweep Sampling of Grasshoppers on Konza Prairie LTER watersheds (Reformatted to ecocomDP Design Pattern)"         
#> [2] "North Temperate Lakes LTER: Pelagic Macroinvertebrate Summary 1983 - current (Reformatted to ecocomDP Design Pattern)"  
#> [3] "North Temperate Lakes LTER: Benthic Macroinvertebrates 1981 - current (Reformatted to ecocomDP Design Pattern)"         
#> [4] "North Temperate Lakes LTER: Pelagic Macroinvertebrate Abundance 1983 - current (Reformatted to ecocomDP Design Pattern)"
```

See `?search_data` for additional query parameters.

## Read

Read data from host APIs to get the newest authoritative version.

Read from the host API:

```{r eval=FALSE}
dataset_1 <- read_data("edi.193.5")
#> Reading edi.193.5
#>  [0%] Downloaded 0 bytes...
#>  [0%] Downloaded 0 bytes...
#>  [0%] Downloaded 0 bytes...
#>  [0%] Downloaded 0 bytes...
#>  [0%] Downloaded 0 bytes...
#>  [0%] Downloaded 0 bytes...
#>  [0%] Downloaded 0 bytes...
#>  [0%] Downloaded 0 bytes...
#>
#> Validating edi.193.5:
#>   Required tables
#>   Column names
#>   Required columns
#>   Column classes
#>   Datetime formats
#>   Primary keys
#>   Composite keys
#>   Referential integrity
#>   Latitude and longitude format
#>   Latitude and longitude range
#>   Elevation
#>   variable_mapping
```

Read from the host API with filters when datasets are large (currently only for NEON datasets):

```{r eval=FALSE}
dataset_2 <- read_data(
  id = "neon.ecocomdp.20120.001.001", 
  site = c("COMO", "LECO", "SUGG"),
  startdate = "2017-06", 
  enddate = "2019-09",
  check.size = FALSE)
#> Finding available files
#>   |==================================================================| 100%
#> 
#> Downloading files totaling approximately 1.588594 MB
#> Downloading 20 files
#>   |====================================================================| 100%
#> 
#> Unpacking zip files using 1 cores.
#>   |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=00s  
#> Stacking operation across a single core.
#> Stacking table inv_fieldData
#>   |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=00s  
#> Stacking table inv_persample
#>   |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=00s  
#> Stacking table inv_pervial
#>   |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=00s  
#> Stacking table inv_taxonomyProcessed
#>   |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=02s  
#> Stacking table inv_taxonomyRaw
#>   |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=02s  
#> Copied the most recent publication of validation file to /stackedFiles
#> Copied the most recent publication of categoricalCodes file to /stackedFiles
#> Copied the most recent publication of variable definition file to /stackedFiles
#> Finished: Stacked 5 data tables and 3 metadata tables!
#> Stacking took 4.732454 secs
#> Joining, by = c("uid", "sampleID")
#> Joining, by = "sampleID"
#>
#> Validating neon.ecocomdp.20120.001.001:
#>   Required tables
#>   Column names
#>   Required columns
#>   Column classes
#>   Datetime formats
#>   Primary keys
#>   Composite keys
#>   Referential integrity
#>   Latitude and longitude format
#>   Latitude and longitude range
#>   Elevation
#>   variable_mapping
```

A data are returned as a list of metadata, tables, and validation issues (if there are any). The dataset identifier is at the top of this list.

```{r include=FALSE}
dataset_1 <- ants_L1
```

```{r}
str(dataset_1)
```

## Manipulate

Working with a "flattened" dataset simplifies common tasks (select, filter, arrange, group, summarize). A "flat" version is where all tables have been joined and spread wide except for the core observation variables, which remain in long form.

```{r}
flat <- flatten_data(dataset_1)
flat
```

## Plot

Visually explore data with the plotting functions:

```{r fig.height=4, fig.width=7}
plot_taxa_accum_time(flat)

plot_taxa_diversity(flat)
plot_taxa_diversity(flat, time_window_size = "month")
plot_taxa_diversity(flat, time_window_size = "year")

plot_sample_space_time(flat)

plot_taxa_shared_sites(flat)

plot_taxa_rank(flat, facet_var = "location_id")

plot_taxa_occur_freq(
  data = flat,
  facet_var = "location_id",
  color_var = "taxon_rank")

plot_taxa_abund(
  data = flat,
  facet_var = "location_id",
  color_var = "taxon_rank",
  trans = "log10")

plot_sites(flat)
```


## Save

Save local copies and read them back in later.

Save a local copy as .rds:

```{r include=FALSE}
datasets <- c(ants_L1, ants_L1)
```

```{r, eval=FALSE}
datasets <- list(dataset_1, dataset_2)
mypath <- paste0(tempdir(), "/data")
dir.create(mypath)

save_data(datasets, mypath)
```

Read a local copy from .rds

```{r, eval=FALSE}
datasets <- read_data(from = paste0(mypath, "/datasets.rds"))
#> Validating edi.193.5:
#>   Required tables
#>   Column names
#>   Required columns
#>   Column classes
#>   Datetime formats
#>   Primary keys
#>   Composite keys
#>   Referential integrity
#>   Latitude and longitude format
#>   Latitude and longitude range
#>   Elevation
#>   variable_mapping
#> Validating neon.ecocomdp.20120.001.001:
#>   Required tables
#>   Column names
#>   Required columns
#>   Column classes
#>   Datetime formats
#>   Primary keys
#>   Composite keys
#>   Referential integrity
#>   Latitude and longitude format
#>   Latitude and longitude range
#>   Elevation
#>   variable_mapping
```

Save a local copy as .csv:

```{r, eval=FALSE}
save_data(datasets, mypath, type = ".csv")
```

Read a local copy from .csv:

```{r, eval=FALSE}
datasets <- read_data(from = mypath)
#> Validating edi.193.5:
#>   Required tables
#>   Column names
#>   Required columns
#>   Column classes
#>   Datetime formats
#>   Primary keys
#>   Composite keys
#>   Referential integrity
#>   Latitude and longitude format
#>   Latitude and longitude range
#>   Elevation
#>   variable_mapping
#> Validating neon.ecocomdp.20120.001.001:
#>   Required tables
#>   Column names
#>   Required columns
#>   Column classes
#>   Datetime formats
#>   Primary keys
#>   Composite keys
#>   Referential integrity
#>   Latitude and longitude format
#>   Latitude and longitude range
#>   Elevation
#>   variable_mapping
```
